---
title: "Choosing a clock"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a clock}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

There is no best clock. There is a clock that answers the question you asked, and a much larger
number that answer a different one in the same units.

## By the question

| Question | Clocks | Scale |
|---|---|---|
| How old does this sample look? | `horvath2013`, `hannum`, `skinandblood`, `lin`, `zhangen`, `zhangblup`, `vidalbralo` | years |
| Who is at risk of dying sooner? | `dnamphenoage`, `hrsinchphenoage`, `zhangmortality` | years, or a log-hazard |
| How fast is this person aging? | `dunedinpoam38` | years per year |
| How old is this newborn, gestationally? | `knight`, `leecontrol`, `leerobust`, `leerefinedrobust` | weeks |
| Is causal or damage-driven aging separable? | `yingcausage`, `yingdamage`, `yingadaptage` | years |
| How dysregulated is this person's physiology? | `phenoage`, `kdm`, `hd` | years, years, distance |

```{r}
list_clocks(generation = "second", tier = "A")
compatible_clocks(d)
```

## Scale governs what you may do

Every clock carries a `scale_type`, and it decides which downstream operations are defined.

| Scale | Legal | Not legal |
|---|---|---|
| `age_years` | acceleration, residual, difference, correlation | - |
| `gestational_weeks` | the same, in weeks | mixing with years |
| `mortality_log_hazard` | correlation, rank, hazard ratio | acceleration: a log-hazard has no zero on the age scale |
| `pace_ratio` | correlation, rank, difference | acceleration: it is already a rate |
| `telomere_kb` | acceleration, correlation, difference | comparison with an age in years |
| `proportion` | compositional operations | anything ignoring the sum-to-one constraint |

`acceleration()` enforces this. Naming a clock whose scale does not support it is an error;
leaving the clock list empty quietly excludes them.

## Coverage is not validity

A clock trained on 450K runs perfectly well on EPIC data that carries its probes, and fails on
450K data filtered to 20,000 probes. Only the feature list can answer whether a clock is usable,
which is what `compatible_clocks()` checks.

But high coverage does not make a result meaningful. The mammalian methylation array was designed
around CpGs conserved across mammals, and carries 96% of Horvath2013's 353 probes - so a zebra
scores at *higher* coverage than many human 450K datasets and returns a confident number from a
clock fitted on people. Nothing in the arithmetic can notice. Set `species` and FALCONAge warns:

```{r}
d <- falcon_data(betas, modality = "dna_methylation", species = "Equus grevyi")
res <- score(d, clocks = "horvath2013")
manifest(res)$warnings
#> trained on Homo sapiens, scored on Equus grevyi. Feature coverage says nothing
#> about whether the coefficients transfer across species.
```

## Availability

| Tier | n | What you do |
|---|---:|---|
| A | 35 | Nothing. Coefficients ship inside the package. |
| B | 98 | Catalogued, but no primary source has been established, so no coefficients ship. Eleven now record where the material is and what blocks it. |
| C | 28 | Obtain a coefficient file and register it. The architecture is implemented and tested. |

```{r}
list_clocks(tier = "C")     # the ones needing author permission
list_clocks(untraced = TRUE)
```

The distinction between B and C matters. C is a licensing decision by the authors. B is an
honesty one by us: copying a coefficient set out of another package without knowing where that
package got it is how the known paper-versus-implementation discrepancies spread in the first
place, and a traced extractor is a welcome contribution.
