---
title: "Getting started with FALCONAge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with FALCONAge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

FALCONAge scores DNA methylation and clinical chemistry against a catalogue of 161 published
aging clocks. The numerical work happens in one Python core that this package calls through
reticulate, so an R result and a Python result are the same bits rather than two implementations
that agree approximately.

## Setup, once

Docker is the supported path. One image carries R, this package, the Python core, the CLI and all
20 bundled clocks at pinned versions, so the same input gives the same numbers on your machine and
on a reviewer's. A plain `sh` fence, since this runs in a terminal rather than in R:

```sh
git clone https://github.com/bhagesh-h/FALCONAge.git && cd FALCONAge
docker build -f docker/Dockerfile.cpu -t falconage:1.1.0-cpu .

# an R session with FALCONAge already installed
docker run --rm -it -v "$PWD:/work" -w /work falconage:1.1.0-cpu R

# or prove the install by running the R suite
docker run --rm -v "$PWD:/work" -w /work falconage:1.1.0-cpu \
  Rscript -e 'testthat::test_local("r")'
```

On Windows PowerShell write `"${PWD}"` in place of `"$PWD"`.

If you would rather install into your own R library, FALCONAge is not on CRAN yet, so both halves
install from GitHub. The R package is in the `r/` subdirectory, hence `subdir`. Again a plain
fence, so `remotes` and `pak` do not become declared dependencies of a package that only mentions
them:

```r
remotes::install_github("bhagesh-h/FALCONAge", subdir = "r")
# or, resolving dependencies better and taking the subdirectory inline:
pak::pak("bhagesh-h/FALCONAge/r")
```

```{r}
library(FALCONAge)
falconage_install()   # creates a managed Python environment; a few minutes
falconage_config()    # what resolved: versions, devices, registry size
```

`falconage_install()` fetches the Python core from the same GitHub tag this package was built
from, so the two halves cannot drift - which matters here more than usual, because the test suite
asserts that R and Python return the same bits, and a mismatched pair fails that assertion in a
way that reads as a numerical bug rather than as version skew. Inside the Docker image the
environment is already built, so `falconage_install()` is not needed there.

Every code chunk below is identical in both, and `falconage_config()` is the first thing to run
when anything looks wrong. It reports the interpreter, the registry version, and how many clocks
are usable:

```
FALCONAge 1.1.0
  registry     1.1.0 (161 clocks)
  tiers        A 23 ship - B 110 need a traced source - C 28 scaffold only
  devices      cpu
```

## Scoring

```{r}
d   <- prepare(read_betas("betas.csv"))
res <- score(d, clocks = "compatible")

summary(res)
as.data.frame(res)                  # samples x clocks
as.data.frame(res, form = "long")   # one row per sample per clock, with the scale
```

`clocks = "compatible"` scores what this dataset can support and reports the rest as skipped with
a reason. Naming clocks explicitly is different: then every one must work, because an explicit
request should never be dropped quietly.

```{r}
coverage(res)     # per-clock feature coverage and skip reasons
manifest(res)     # versions, device, and the SHA-256 of every coefficient file used
```

## Age acceleration, and which convention you mean

The three conventions disagree by several years on the same data, and papers frequently do not
say which they used.

```{r}
acceleration(res, method = "absolute")      # predicted - chronological
acceleration(res, method = "residual")      # residual of predicted on chronological
acceleration(res, method = "within_group", group = "dataset")
```

Absolute is interpretable in years and carries the clock's own bias: a clock five years high on
everybody gives everybody five years of acceleration. Residual removes that bias, and removes any
real cohort-wide effect with it. Within-group is what the AA2 benchmark needs - it asks whether
cases accelerate relative to *their own* controls.

Ask for acceleration on a pace-of-aging clock and it refuses:

```{r}
acceleration(res, clocks = "dunedinpoam38")
#> Error: 'acceleration' is not defined for dunedinpoam38, whose output is
#>   pace_ratio (biological years per chronological year).
#>   A pace of aging is already a rate; subtracting chronological age from it is
#>   a units error, not a conservative choice.
```

That refusal is the point. A wrong number in a column of numbers is invisible.

## The clock catalogue

```{r}
list_clocks(tier = "A")             # ship with coefficients, run offline
list_clocks(search = "mortality")
clock_info("horvath2013")
cite_clock("dnamphenoage", "bibtex")
```

Twenty-eight clocks are scaffolds: the model, the feature list, the transform chain and the
expected shapes all ship, but the coefficients are research-use-only and are not ours to
distribute. `clock_info()` on one says why, where to obtain a file, and which open clocks answer
the same question.

```{r}
register_local_weights("grimage2", "~/licensed/grimage2_coefs.csv")
score(d, clocks = "grimage2")
```

## Clinical chemistry, and units

```{r}
d <- read_clinical("nhanes.csv", units = list(
  albumin = "g/L", creatinine = "umol/L", glucose = "mmol/L", crp = "mg/dL",
  lymphocyte_percent = "%", mean_cell_volume = "fL",
  red_cell_distribution_width = "%", alkaline_phosphatase = "U/L",
  white_blood_cell_count = "10^3/uL", age = "years"))

score(d, clocks = "phenoage")
```

Units are required and never guessed. Albumin at 4.2 is g/dL and albumin at 42 is g/L; both are
clinically normal, and PhenoAge fitted on one returns nonsense for the other. No range check can
separate them, so the only correct behaviour is to ask.

KDM and homeostatic dysregulation have no fixed coefficients at all - they are defined relative
to a reference cohort:

```{r}
ref <- fit_kdm(nhanes3, markers = markers)
score(d, clocks = "kdm", reference = ref)
```

## Benchmarking

```{r}
res <- combine(list(r1, r2, r3), keys = c("GSE1", "GSE2", "GSE3"))
b   <- run_benchmark(res, condition_col = "condition", control = "HC",
                     dataset_col = "dataset")
b$summary
```

Median absolute error against chronological age is reported but never ranked on: a perfect
chronological oracle would score best on it and be useless, because it would have no age
acceleration left to detect anything with.

## Reporting

```{r}
write_results(res, "results/")
report(res, "report.html", group = "condition")
```
